---
title: "Plate"
author: "Kai Aragaki"
date: "2021-08-10"
output: tufte::tufte_html
---

```{r, echo=FALSE, message=FALSE, results='hide'}
library(methods)
library(purrr)
library(dplyr)


setClass("Plate",
         slots = c(
           layers = "list",
           meta = "list",
           wells = "numeric"),
         prototype = c(
           layers = list(),
           date = list(),
           wells = NA_real_
         )
)

setValidity("Plate", function(object) {

  # Ensure wells is a valid number of wells
  if (!is.na(object@wells) && !(object@wells %in% c(6, 12, 24, 48, 96, 384, 1536, 3456))) {
    return("`wells` does not match a valid form factor (6, 12, 24, 48, 96, 384, 1536, or 3456)")
  }

  # If the wells slot it set, enforce incoming layer dimensions


  # Each layer must have dimensions
  any_null_dims <- object@layers |>
    lapply(dim) |>
    lapply(is.null) |>
    unlist() |>
    any()

  if (any_null_dims) {
    return("All layers must have dimensions")
  }


  if (length(object@layers) < 1) {
    TRUE
  } else {
    unique_row_dims <- object@layers |>
      lapply(nrow) |>
      unlist() |>
      unique() |>
      length()

    unique_col_dims <- object@layers |>
      lapply(ncol) |>
      unlist() |>
      unique() |>
      length()

    if (!(unique_row_dims == 1 & unique_col_dims == 1)) {
      "All layers must have the same dimensions"
    } else {
      TRUE
    }
  }
})

Plate <- function(layers = list(), wells = NA_real_, meta = list()) {
  new("Plate", layers = layers, wells = wells, meta = meta)
}



setGeneric("serve", function(x) standardGeneric("serve"))

setMethod("serve", "Plate", function(x) {
  x@layers |>
    lapply(as.matrix) |>
    lapply(as.vector) |>
    reduce(bind_cols) |>
    setNames(names(x@layers))
})


# Layer(s) getter ----
setGeneric("spill", function(x, ...) standardGeneric("spill"))
setMethod("spill", "Plate", function(x, n = NULL) {
  if (is.null(n)) {
    x@layers
  } else {
    x@layers[n]
  }
})

setGeneric("layers", function(x, ...) standardGeneric("layers"))
setMethod("layers", "Plate", function(x, n = NULL) {
  if (is.null(n)) {
    x@layers
  } else {
    x@layers[n]
  }
})

# Layer setter ----
setGeneric("layers<-", function(x, ...) standardGeneric("layers<-"))
setMethod("layers<-", "Plate", function(x, value, position = NULL) {
  if(is.null(position)) {
    x@layers[[length(salad@layers) + 1]] <- value
  } else {
    x@layers[[position]] <- value
  }
  validObject(x)
  x
})


# Well getter ----
setGeneric("wells", function(x) standardGeneric("wells"))
setMethod("wells", "Plate", function(x) {
  x@wells
})

# Well setter ----
setGeneric("wells<-", function(x, ...) standardGeneric("wells<-"))
setMethod("wells<-", "Plate", function(x, value) {
  x@wells <- value
  validObject(x)
  x
})
set.seed(1)
```

# Overview

Micro-well plates are incredibly common tools used in the lab. Their rectangular nature often causes the data they yield to be rectangular. One example is a protein quantification assay, which produces a rectangular array of numbers associated with each wells absorbance at 562nm.

Despite being rectangular, these data are NOT tidy: a column does not refer to a variable, and row does not refer to a case, unless you are very specific about your plating strategies. Therefore, working with these data is a computational manner is somewhat difficult for almost all modern workflows, which expect tidy data. 

However, a simple conversion from 'plate data' to tidy data loses an import aspect of the physical nature of the plate. Plates often store data spatially, due to sample layout. Because of this, data entry about sample location in a tidy format is cumbersome.

`Plate` provides a set of tools to work with plate data, as well as methods to move between 'tidy' and 'plate' data. Additionally, it provides a framework that other scripts and packages can build upon to work with plate data.

# In action

Plate construction is incredibly simple. Plates take a list of matrices or matrix-like objects (`data.frame`s, for instance) and store them as 'layers':

```{r}
leaves <- matrix(sample(1:10, 6, replace = TRUE), 2, 3)
tomatoes <- matrix(sample(0:1, 6, replace = TRUE), 2, 3)
portions <- matrix(c("max", "jerry", "david", "larry", "beth", "judy"), 2, 3)

salad <- Plate(layers = list(leaves = leaves,
                             tomatoes = tomatoes,
                             portions = portions))
salad
```

The number of wells can be stored, also:
(This well slot will in the future be used to validate incoming data to make sure it has the correct dimensions)

```{r}
wells(salad) <- 6
salad
```

And will complain if you don't do it right:

```{r error=TRUE}
wells(salad) <- "a lot"
```

```{r error=TRUE}
wells(salad) <- 7
```

Plates can have layers added to them:

```{r}
layers(salad) <- data.frame(left = c(1, 2),
                            middle = c(2, 1),
                            right = c(1, 2))
```

If you don't want to deal with all the trappings of 'Plate-ness', you can `spill` the data off the `Plate` to get a list of arrays:

```{r}
spill(salad)
```

or spill individual layers:

```{r}
spill(salad, "leaves")
```


Perhaps most importantly: you can `serve` `Plate`s tidily:

```{r}
serve(salad)
```

